Variety in the Implementation of Nonlinear Least Squares Program Codes

### 2022-06-15 VarietyInNLS/VarietyInNonlinearLeastSquaresCodes.Rmd
Begin simplifying and shortening Variety

There is no exogenous data here.

There may be additional conditions that add to the complexity of
the problem such as

@@

An overview of the method is provided by Wikipedia: *https://en.wikipedia.org/wiki/Gauss%E2%80%93Newton_algorithm*.

The method was described by C.F.Gauss in 1809 in *Theoria motus corporum 
coelestium in sectionibus conicis solem ambientum*.

@@

$$ x \leftarrow x + \delta $$

and we iterate until there is no change in $x$, assuming that the method does indeed
converge for the problem at hand. Newton (and his sometimes associated worker Raphson) 
used the algorithm we attribute
to them in a very special case and in a manner unrecognizable as that described
above.

@@

$$  J^t J $$

Arguing that the residuals should be "small", Gauss proposed that the 
Newton equations could be approximated by ignoring
the second term (which has elements of the residual vector times second derivatives 
of the residual).  This gives the **Gauss-Newton**
equations (cancelling a factor 2)

$$ (J^t J)  \delta  =  - J^t r $$

We can use this in an iteration similar to the Newton iteration. Note that these
equations are equivalent to the linear least squares problem that can be written

$$  J \delta  \approx - r $$

Like most other treatments, we will skate over the thin ice that $J^t J$ is a 
good approximation of the Hessian and that the resulting method converges.

@@

$$ x \leftarrow x + \delta$$

one could do a search along the direction $\delta$. Ideally we would want to **minimize**

$$ f(x + \alpha \delta) $$

with respect to the (scalar) parameter $\alpha$. This one-dimensional minimization can
take quite a lot of computational effort, however, and typically a value of $\alpha$ that
permits a suitable reduction in the sum of squares $f(x)$ is accepted, and the iteration repeated
from

$$ x_{new} = x + \alpha * \delta $$

Clearly there are many tactical choices that give rise to a variety of particular algorithms.
One concern is that the direction $\delta$ may give no lower value of the sum of squares, since
there is an approximation involved in using $J^t J$ rather than $H$. For example, $\delta$ could
be perpendicular to any downhill direction.

@@

In `nls.c` we also see that once a satisfactory "new" set of parameters has been 
obtained, we reset the $\alpha$ value to the minimum of $2 * \alpha$ and 1.0. Oddly, there
is no provision for the user to set the tactical choices of the maximum value, the reduction
or the increase in $\alpha$, despite our experience that other choices can be advantageous
for some problems. Moreover, $x_{new}$ is deemed satisfactory if its **deviance** (essentially
the weighted sum of squares of the residuals) is **no greater than** the current best
value. Our experience is that it may be better to insist on a decrease in the value,
or even to use an "acceptable point" approach that enforces a sufficient decrease using
a measure based on the gradient of the deviance projected on $\delta$.

@@

$$  K \delta \approx - r_{augmented} $$

where we have added $n$ zeros to $r$ to make $r_{augmented}$. Similar adjustments
are used for scaled Marquardt methods mentioned below.

@@

@@

While the above ideas are the dominant theme in nonlinear least squares, there have been many 
other proposed approaches. The following is a non-exhaustive sample.

### Spiral

@Jones70Spiral seems only to be mentioned in a single paper. There is a flowchart
given but I have never been able to find a code (supposedly Fortran V for the Univac
1108.) The motivations are to avoid both matrix equation solutions and one-dimensional
line-search minimization. Jones mentions matrix inversion, but it is not clear whether
he truly forms the inverse. The main idea in the paper (and presumably the code) is
to use a geometric spiral that mimics the Marquadt search trajectory as $\lambda$
is increased. It would be interesting to actually try some of the ideas, which do,
as far as I can determine, have a reasonable basis and could possibly offer some
efficiencies.

### Approaches to Newton's method for special models

Newton (and his sometime associated worker Raphson) used the algorithm we attribute
to them applied to a very special case and in a manner unrecognizable as that described
above.  (See *https://en.wikipedia.org/wiki/Newton%27s_method*).

The main attraction of the Gauss-Newton method is that the Jacobian only needs first
derivatives, which are generally much, much easier to compute than the full second
derivatives of the Hessian. However, there have been suggestioned
Newton-like approaches, possibly for special types of models where some short cut
to the Hessian is available. These may or may not make sense in the context of
a given user or group of users. I regard these approaches as part of the
computational infrastructure of the special cases rather than as general nonlinear
least squares methods. The `maxLik` package (@maxLik2011) is a prime example.

### Hybrid methods

The central idea of Hartley's method is a line search and that of Marquardt a
parametrized stabilization that can be shown to reduce a sum of squares for a large
enough stabilization parameter. There is nothing to prevent use of a line
search additional to the Marquardt stabilization, and there are, of course, 
several reasonable choices for line search. Also we could solve the (possibly
stabilized) Gauss-Newton equations by several quite different numerical linear
algebra techniques, or even approximations. In certain situations, some of these
choices may be important. Unfortunately, the effort to benchmark the possibilities
seems higher than any benefit in replacing quite good existing methods.

@@

# Implementation comparisons

@@

>>>>

There is no exogenous data here. However, there may be additional 
conditions that add to the complexity of the problem such as

@@

An overview of the method is provided by Wikipedia: *https://en.wikipedia.org/wiki/Gauss%E2%80%93Newton_algorithm*.
Indeed, the method was described by C.F.Gauss in 1809 in *Theoria motus corporum 
coelestium in sectionibus conicis solem ambientum*.

@@

$$ x \leftarrow x + \delta $$
and we iterate until there is no change in $x$, assuming that the method does indeed
converge for the problem at hand. Newton (and his sometimes associated worker Raphson) 
used the algorithm we attribute
to them in a very special case and in a manner unrecognizable as that described
above.

@@

$$  J^t J $$
Arguing that the residuals should be "small", Gauss proposed that the 
Newton equations could be approximated by ignoring
the second term (which has elements of the residual vector times second derivatives 
of the residual).  This gives the **Gauss-Newton**
equations (cancelling a factor 2)

$$ (J^t J)  \delta  =  - J^t r $$
We can use this in an iteration similar to the Newton iteration. Note that these
equations are equivalent to the linear least squares problem that can be written

$$  J \delta  \approx - r $$
Like most other treatments, we will skate over the thin ice that $J^t J$ is a 
good approximation of the Hessian and that the resulting method converges.

@@

$$ x \leftarrow x + \delta$$
one could do a search along the direction $\delta$. Ideally we would want to **minimize**

$$ f(x + \alpha \delta) $$
with respect to the (scalar) parameter $\alpha$. This one-dimensional minimization can
take quite a lot of computational effort, however, and typically a value of $\alpha$ that
permits a suitable reduction in the sum of squares $f(x)$ is accepted, and the 
iteration repeated from

$$ x_{new} = x + \alpha * \delta $$
Clearly there are many tactical choices that give rise to a variety of particular algorithms.
One concern is that the direction $\delta$ may give no lower value of the sum of squares, 
since there is an approximation involved in using $J^t J$ rather than $H$. 
For example, $\delta$ could be perpendicular to any downhill direction.

@@

In R's `nls.c` we also see that once a satisfactory "new" set of parameters has been 
obtained, we reset the $\alpha$ value to the minimum of $2 * \alpha$ and 1.0. Oddly, there
is no provision for the user to set the tactical choices of the maximum value, the reduction
or the increase in $\alpha$, despite our experience that other choices can be advantageous
for some problems. Moreover, $x_{new}$ is deemed satisfactory if its **deviance** (essentially
the weighted sum of squares of the residuals) is **no greater than** the current best
value. Our experience is that it may be better to insist on a decrease in the value,
or even to use an "acceptable point" approach that enforces a sufficient decrease using
a measure based on the gradient of the deviance projected on $\delta$.

@@

$$  K \delta \approx - r_{augmented} $$
where we have added $n$ zeros to $r$ to make $r_{augmented}$. Similar adjustments
are used for scaled Marquardt methods mentioned below.

@@

<!-- ### Variants -->

<!-- The central idea of Hartley's method is a line search and that of Marquardt a -->
<!-- parametrized stabilization that can be shown to reduce a sum of squares for a large -->
<!-- enough stabilization parameter. There is nothing to prevent use of a line -->
<!-- search additional to the Marquardt stabilization, and there are, of course,  -->
<!-- several reasonable choices for line search. Also we could solve the (possibly -->
<!-- stabilized) Gauss-Newton equations by several quite different numerical linear -->
<!-- algebra techniques, or even approximations. In certain situations, some of these -->
<!-- choices may be important. Unfortunately, the effort to benchmark the possibilities -->
<!-- seems higher than any benefit in replacing quite good existing methods. -->

@@

While the above ideas are the dominant themes in nonlinear least squares, 
there have been many other approaches. 
<!-- The following is a non-exhaustive sample. -->

<!-- ### Spiral  -->

<!-- @Jones70Spiral seems only to be mentioned in a single paper. There is a flowchart -->
<!-- given but I have never been able to find a code (supposedly Fortran V for the Univac -->
<!-- 1108.) The motivations are to avoid both matrix equation solutions and one-dimensional -->
<!-- line-search minimization. Jones mentions matrix inversion, but it is not clear whether -->
<!-- he truly forms the inverse. The main idea in the paper (and presumably the code) is -->
<!-- to use a geometric spiral that mimics the Marquadt search trajectory as $\lambda$ -->
<!-- is increased. It would be interesting to actually try some of the ideas, which do, -->
<!-- as far as I can determine, have a reasonable basis and could possibly offer some -->
<!-- efficiencies. -->

### Using Newton's method for special models

Newton (and his sometime associated worker Raphson) used the algorithm we attribute
to them applied to a very special case and in a manner unrecognizable as that described
above.  (See *https://en.wikipedia.org/wiki/Newton%27s_method*). Since Newton's
method requires the full Hessian and its second derivatives, we do not see it applied
except when the Hessian is easily and efficiently calculated.
The Gauss-Newton method needs only first derivatives for the Jacobian. Generally, 
these are much easier to compute. However, for special types of models where some short cut
to the Hessian is available, Newton may be worthwhile. The `maxLik` package (@maxLik2011) 
for R is an example.

### Using general optimization methods

There is no prohibition on applying algorithms for general function minimization, 
even with constraints, to problems where the objective is a sum of squared functions.
General methods also contend with issues of approximating the Hessian
or ensuring iterates are satisfactory, but there may be additional advantages
with a sum of squared function.

?? nlminb (from Gay etc. nlr)
?? nl2sol (??)
?? DUD ??

# Differences in user interface

@@

# Implementation considerations

@@

?? More 77 item
